Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-10-12 doc/JFLA2022/paper.tex
complete pass over section 3

\subsection{Hydra games}

We chose to build our library and book on two simple themes which allow many variations: computing powers in a monoid, and Kirby and Paris' hydra battles. In the interest of space, we shall present only the second one in this article.

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionally presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:round}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater than 1 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
  Figure~\ref{fig:round} shows an example with $n=2$.
\end{itemize}

@@

\section{Integration within \community}

\subsection{Presentation}

\community is an informal organization run by volunteer Coq users with objectives to maintain interesting Coq projects in the long-term and help Coq users to collaborate on documentation, tooling, etc.

It was created in 2018, inspired by the {Elm Community} organization~\cite{zimmermann:tel-02451322}.
%
Such ``Community Package Maintenance Organizations'' actually exist in many ecosystems as they avoid the very common problem of an important package becoming unmaintained as its author has moved on or disappeared~\cite{zimmermann2021grounded}.

In the case of \coq, this problem may be even more prevalent as many packages are created by PhD students, or by researchers for a specific paper, but their authors do not intend to maintain the package in the long-term.
%
The authors are however generally very open to having someone else who expresses interest in their work continue the maintenance.
%

\community makes that easy by providing a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration) and by making it possible for someone to take over a package without committing in the long-term (as maintainers who drop out can easily be replaced by some other volunteers).

Today, \community hosts over 50 projects maintained by over 30 maintainers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the \coq development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new \coq versions were conducted.
%
Some were still maintained by their authors but they wanted help to keep them up-to-date.
%
Others were simply left unmaintained and were revived by their adoption in \community.

@@

Given the objectives of \community, it makes sense to propose a transfer anytime we encounter an interesting project that is insufficiently maintained.
%
After the project is adopted, large changes and refactorings are explicitly allowed.
%
This means, for instance, that we can consolidate libraries together or split them into several packages if it makes sense.

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem~\cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \community.

Computability is a key topic in Computer Science teaching. Moreover O'Connor's library is a nice illustration of dependently typed programming, so we chose to devote a full chapter (chapter 9 of~\cite{HydraBook}) to this formalization, with comments on the definitions and proofs, [counter]-examples and exercises, and  extract part of the Goedel library into a new sub-library of the Hydra-battles library: Ackermann.

>>>>

\subsection{Hydra games}
Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionally presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:round}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater than 1 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
  Figure~\ref{fig:round} shows an example with $n=2$.
\end{itemize}

@@

\section{Integration with \community}

\community is an informal organization run by volunteer Coq users on GitHub that aims to maintain interesting Coq projects for the long term and facilitate collaboration among Coq users on documentation, tooling, etc. \community was created in 2018, inspired by the {Elm Community} organization~\cite{zimmermann:tel-02451322}.
%
Such ``Community Package Maintenance Organizations'' exist in many software ecosystems, as they avoid the common problem of an important package becoming unmaintained after its author has moved on to other projects or has disappeared~\cite{zimmermann2021grounded}.

In the case of \coq, this problem is likely even more prevalent than in other ecosystems, as many packages are created by graduate students or researchers for a specific paper and not planned to be maintained for the long term by authors.
%
However, authors are generally open to having someone else who expresses interest in their work take over package maintenance.
%

\community makes it easy to change maintainers by defining a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration), and by making it possible for someone to take over a package without making a long-term commitment (as maintainers who drop out can easily be replaced by some other volunteers).

As of 2021, \community hosts over 50 projects maintained by over 30 volunteers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the \coq development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new \coq versions were performed.
%
Some were still maintained by their original authors, but were transferred to enable other users to help out with maintenance and facilitate adoption of best practices on, e.g., continuous integration.
%
Others were simply unmaintained and were revived after their transfer to \community.

@@

Given the objectives of \community, we believe it makes sense to propose a transfer anytime we encounter an interesting Coq project that is insufficiently maintained.
%
After a transfer, the \community maintainers are explicitly allowed to perform large changes and refactorings.
%
This means, for instance, that we can consolidate libraries by merging them, or split them up into several packages.

In order to prove formally that the length of the
kind of hydra battles we consider is not given by a primitive recursive function, we chose to use a formalization of the primitive recursive functions that was originally part of Russell O'Connor's statement and proof of G\"{o}del's first incompleteness theorem~\cite{OConnor05, Goedel}.
For this purpose, and above all in consideration of the scientific interest of this contribution, we decided to host and maintain O'Connor's work in \community.

Since computability is a key topic in computer science teaching and O'Connor's library is a nice illustration of dependently typed programming, we decided to devote a full chapter (chapter 9 of~\cite{HydraBook}) to the formalization of primitive recursive functions, with comments on the definitions and proofs and [counter-]examples and exercises. As part of the writing process, we made the formalization into a new sub-library of the Hydra-battles library, dubbed Ackermann.
